---
title: "Class 05 Data Visualization with GGPLOT"
author: "Chloe"
format: pdf
---

# Our first ggplot

To use the ggplot2 package I first need to have it installed on my computer

To install my package we use `install.packages()` command.

Now can I use it? NO! First we need to call `library(ggplot2)`

```{r}
library(ggplot2)
ggplot()
```

```{r}
mpg
```


Our first plot of displ vs hwy
All ggplot() graphs are made in the same way
- data + aes + geoms

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) + 
  geom_point()
```

I can add more layers:

```{r}
ggplot(mpg) +
  aes(x=displ, y=hwy) + 
  geom_point()+
  geom_smooth(method = lm, se = FALSE)
```

# Plot of gene expression data

First read the data from online.

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)

```
> Q. How many genes are in this dataset?

```{r}
nrow(genes)
```


What are the colnames?
```{r}
colnames(genes)
```

A first version plot of this data Condition1 vs Condition2

```{r}
ggplot(genes)+ 
  aes(x=Condition1, y=Condition2) + 
  geom_point()

```

```{r}
p <- ggplot(genes)+ 
  aes(x=Condition1, y=Condition2, col=State) + 
  geom_point()
p
```

> Q. How many genes are up regulated and down regulated?

```{r}
head(genes)
```

To get at just the State column
```{r}
table(genes$State)

```

```{r}
round( table(genes$State)/nrow(genes) * 100, 2 )
```

```{r}
p + scale_colour_manual( values=c("blue","gray","red")) +
  labs(title = "Gene Expression Changes Upon Drug Treatment",
       x="Control (no drug)", y="Drug Treatment",
       caption = "Made by Chloe")

```



